---
output: 
  pdf_document: 
    fig_height: 3
    fig_crop: false
fontsize: 10pt
geometry: margin=0.6in
---
## Unit 07 Homework Q3.4
```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

```{r load data, include=FALSE}
data_file_path <- 'spring_2024_central\\data\\GSS_religion.csv'
data <- read.csv(data_file_path)
summary(data)
```
### Attitudes toward the religious
The current recommendation is to apply paired t-test to test this statement. For this test to be valid, it must consider the following assumptions:  
1. The groups are paired  
2. The values are on a metric scale  
3. The data is independent and identically distributed  
4. The distribution of the difference between measurements has no major deviations from normality, considering the sample size.  

Or first assumption is that our groups are paired. Our observation groups are generated from asking multiple questions to the same individual. 

Our second assumption is that our variables are on a metric scale. To test this assumption we can look at a summary of our values. The values for feelings toward Protestants is as follows:
```{r summary of protestant, echo=FALSE}
summary(data$prottemp)
```
The values for feelings toward Protestants is as follows:
```{r summary of catholic, echo=FALSE}
summary(data$cathtemp)
```
These are generated from a question about "feeling" but are on a scale of 0 to 100 and the respondent is prompted to equate this to temperature.  

Our third assumption is that the data is independent and identically distributed. There is risk that some respondents are from the same household and share similiar values and views.   

The distribution of the difference between measurements has no major deviations from normality, considering the sample size. We can create a histogram each of the two variable groups to test this:  

```{r differences, echo=FALSE}
data$diffs = data$prottemp - data$cathtemp

ggplot() +
    geom_histogram(aes(data$diffs), binwidth = 5, color = "grey40", fill = "royalblue4") + 
        labs(title = "Distribution of Differences: Protestant - Catholic Temps
(Dashline = mean)", x = "Difference of Protestant - Catholic Temps", y = "Frequency") +
    geom_vline(aes(xintercept=mean(data$diffs)), linewidth=1, linetype = "longdash")
```
Additionally, we can look at the mean and median of the difference values:
```{r, echo=FALSE}
summary(data$diffs)
```